---
title: "Detecting SMS Spam Using Machine Learning and Deep Learning Approaches"
subtitle: "MH4510 Group 7: Pattern Miners"
author: "Adele Teng(U2240156B), Clarrance Seah Jing Xiong(U2140255A), Wong Zi Xun(U2240117A), Kevin Lee Zhi Hong(U2140142L)"


output:
  beamer_presentation:
    latex_engine: xelatex
    colortheme: lily
    fig_caption: no
    fonttheme: structurebold
    slide_level: 2
    theme: Darmstadt
    toc: no
  slidy_presentation: default
  ioslides_presentation: default
classoption: t
header-includes: 
  - \titlegraphic{\centering \includegraphics[height=2cm]{ntu_logo.png}}
  - \usepackage{animate}
  - \usepackage{booktabs}
  - \setbeamercolor{block title}{bg=blue!75!black,fg=white}
  - \setbeamercolor{block body}{bg=white!90!blue}
fontsize: 9pt
always_allow_html: yes

---

```{r r global_options, include=FALSE, message=FALSE}
# the following command prevents all the R codes 
# from being included into the slides
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE,
                      out.width = "100%", out.height = "70%",
                      fig.align='center', fig.width=10, fig.height=7 )

library(tidyverse)
library(tidytext)
library(caret)
library(ggplot2)
library(dplyr)
library(textstem)
library(tm)
library(wordcloud)

library(stargazer) # for printing nice regression tables

```

```{r}
sms_data <- read.csv("spam.csv")


# Remove duplicates
sms_data <- sms_data[!duplicated(sms_data), ]


sms_data <- sms_data %>% 
  select(v2,v1) %>%
  rename(text = v2, label = v1) %>%
  mutate(label = factor(label, levels = c("ham", "spam"), labels = c(0,1)))


data<- sms_data %>%
  mutate(text = text %>% 
           str_replace_all("[^A-Za-z@]", " ") %>%
           str_squish() %>%
           str_to_lower() %>%
           str_replace_all("@", "twacc")) %>% # emails will use @
  mutate(text= lemmatize_strings(text)) %>%
  mutate(text = removeWords(text, stopwords("en")))
```
```{r}

# Plotting word cloud for ham and spam data 

ham_data <- data[data$label == 0,]
spam_data <- data[data$label == 1,]

ham_words_freq <- termFreq(ham_data$text)
spam_words_freq <- termFreq(spam_data$text)
```
## Introduction

\begin{block}{Background}
  \begin{itemize}
    \item Spam messages are a growing cybersecurity and privacy concern, SMS spam affects billions of users and businesses annually.
    \item Detecting spam prevents phishing, scams, and data theft.
    \item A classic but still evolving NLP challenge.
    \item Manual filtering is inefficient and automation via text mining and ML is essential.
  \end{itemize}
\end{block}

\begin{block}{Objective} 
Develop and compare models that automatically classify SMS as ham (0) or spam (1).
\end{block}

\begin{block}{Dataset}
  \begin{itemize}
    \item \href{https://www.kaggle.com/datasets/uciml/sms-spam-collection-dataset}{\color{blue}Public SMS Spam Collection Dataset (UCI/Kaggle)}
    \item 5,169 unique messages after cleaning
    \item \textasciitilde{}88\% ham, 12\% spam
  \end{itemize}
\end{block}

## Data preprocessing & Exploratory Data Analysis
\begin{columns}
    \begin{column}{0.3\textwidth}
    \begin{block}{Data Cleaning}
    \fontsize{6pt}{7pt}
    \begin{enumerate}
      \item Removed duplicates and nulls.
      \item Converted text to lowercase.
      \item Removed non-alphabetic symbols, lemmatized words.
      \item Replaced “@” with placeholder token “twacc”.
      \item Removed English stopwords.
    \end{enumerate}
  \end{block}
\end{column}

\begin{column}{0.3\textwidth}
  \begin{block}{Feature Engineering}
  \fontsize{6pt}{7pt}
    \begin{itemize}
      \item Bag-of-Words (DTM)
      \item TF-IDF weighting
      \item Word Embeddings (GloVe, 50–100 dimensions)
    \end{itemize}
  \end{block}
```{r bar_distribution, out.height="20%"}
## Data set distribution of classes
ggplot(data=data, aes(x=label, fill=label)) + geom_bar()
```
\end{column}

\begin{column}{0.4\textwidth}
```{r freq_words, out.width="105%", out.height="40%"}
# Comparing frequency of most common words in both labels
library(ggplot2)
library(dplyr)
library(tidytext)

top_words <- data %>%
  unnest_tokens(word, text) %>%
  count(label, word, sort = TRUE) %>%
  group_by(label) %>%
  slice_max(n, n = 30) %>%
  ungroup() %>%
  mutate(word = reorder_within(word, n, label))

ggplot(top_words, aes(x = n, y = word, fill = factor(label))) +
  geom_col(show.legend = FALSE) +
  facet_wrap(~ label, scales = "free_y") +
  scale_y_reordered() +
  labs(
    title = "30 most common Words in each label",
    x = "Frequency",
    y = NULL
  ) +
  theme_minimal() +
  theme(axis.text.y = element_text(size = 8))

# There is quite a bit of overlap, such as 'u', 'call', 'now', 'get', 'just'. Not as much overlap as wanted, but at least that means the word-frequency alone isnt rly enough to predict the labels. 
```
\end{column}
\end{columns}


\begin{columns}
  \begin{column}{0.4\textwidth}
  \begin{block}{Key Insights (EDA)}
    \fontsize{6pt}{7pt}
    \begin{itemize}
      \item Spam words: “free”, “win”, “call”, “claim”, “txt”.
      \item Ham words: “ok”, “thanks”, “see”, “tomorrow”.
      \item Strong class imbalance (ham >> spam).
    \end{itemize}
  \end{block}
\end{column}

\begin{column}{0.3\textwidth}
\begin{figure}
\centering
\includegraphics[width = 0.75\textwidth]{ham_word_cloud.png}
 \end{figure} 
\end{column}

\begin{column}{0.3\textwidth}
\begin{figure}
\centering
\includegraphics[width = 0.75\textwidth]{spam_word_cloud.png}
 \end{figure} 
\end{column}
\end{columns}



\begin{columns}
\begin{column}{1.05\textwidth}
  \fontsize{8pt}{9pt}\selectfont
  \textbf{Interpretation:} EDA shows distinct linguistic patterns: spam messages contain incentive-related and promotional terms, while ham messages are conversational.
\end{column}
\end{columns}


## Model Performance
\begin{columns}
\begin{column}{1.3\textwidth}
\textbf{Model Overview}
```{r model_tbl, results = "asis",include=TRUE, out.height="200%"}
library(kableExtra)
library(tidyverse)


data.frame(Model = c("XGBoost", 
            "BoW+TF-IDF+LASSO", 
            "GloVe+LASSO", 
            "GloVe+LSTM"),
  Representation = c("BoW", 
                     "BoW+TF-IDF", 
                     "50-dim embeddings", 
                     "100-dim Embeddings"),
  Method = c("Ensemble Trees", 
             "L1-Regularized \nLogistic Regression", 
             "Sparse Logistic Regression", 
             "Recurrent Neural Network"),
  "Highlights/Parameters" = c(
    "Logistic objective \nEarly stopping \n100 rounds",
    "Cross-validation for \n$\\lambda$ selection",
    "Semantic-aware \nrepresentation",
    "50 LSTM units, \nDropout 0.2, \n50 epochs"
  )) %>%
  mutate_all(linebreak) %>%
  kable(format='latex',
        col.names = c('Model',"Representation","Method","Highlights/Parameters"), booktabs = T, escape = FALSE) %>% 
  kable_styling(latex_options = c("HOLD_position"), font_size = 7) %>% 
  row_spec(0,bold=T,color = 'black', background = 'white')

```

\end{column}
\end{columns}

\vspace{0.5cm}
\begin{columns}
\begin{column}{1.3\textwidth}
\textbf{Performance Summary}
```{r perf_tbl, results = "asis",include=TRUE, out.height="200%"}
library(kableExtra)
library(tidyverse)

data.frame(Model = c("XGBoost", 
            "BoW+TF-IDF+LASSO", 
            "GloVe+LASSO", 
            "GloVe+LSTM"),
  Accuracy  = c("0.973", 
                "0.961",
                "0.935",
                "0.951"),
  AUC  = c("0.977", 
           "0.965",
           "0.940",
           "0.932"),
  Comments  = c(
    "Strong accuracy, interpretable", 
    "Sparse and interpretable",
    "Semantic-rich, lightweight",
    "Semantic-rich, slow to train")
  )%>%
  mutate_all(linebreak) %>%
  kable(format='latex',
        col.names = c('Model',"Accuracy","AUC","Comments"), booktabs = T, escape = FALSE) %>% 
  kable_styling(latex_options = c("HOLD_position"), font_size = 7) %>% 
  row_spec(0,bold=T,color = 'black', background = 'white')
```

\end{column}
\end{columns}


## Key Takeaways
\begin{columns}
\begin{column}{0.5\textwidth}
  \begin{block}{Key Findings}
    \fontsize{7pt}{8pt}
    \begin{itemize}
      \item Best model: XGBoost (Accuracy 97.3\%)
      \item Best interpretable model: TF-IDF + LASSO (Accuracy 96.1\%)
      \item Most efficient model: XGBoost (Accuracy 97.3\%)
    \end{itemize}
  \end{block}
  
  \begin{block}{Limitations}
    \fontsize{8pt}{9pt}
    \begin{itemize}
      \item Class imbalance could bias model toward ham.
      \item LSTM’s sequential input was averaged — limits temporal learning.
      \item Generalization to multilingual or new spam sources not tested.
    \end{itemize}
    \end{block}
  
\end{column}

\begin{column}{0.5\textwidth}
  \begin{block}{Interpretability}
    \fontsize{8pt}{9pt}
    \begin{itemize}
      \item LASSO coefficients highlight strong predictors: “free”, “win”, “txt”, “call now”.
      \item XGBoost feature importance confirms similar signals.
    \end{itemize}
  \end{block}
  
  \begin{block}{Future Work}
    \fontsize{8pt}{9pt}
    \begin{enumerate}
      \item Integrate contextual embeddings (BERT).
      \item Use SMOTE or cost-sensitive learning for imbalance correction.
      \item Evaluate cross-dataset transferability.
    \end{enumerate}
  \end{block}
\end{column}
\end{columns}

\vspace{0.3cm}
\begin{columns}
\begin{column}{1.05\textwidth}
  Combining strong preprocessing, interpretable models, and deep learning leads to robust spam detection systems that can be deployed in real-world messaging platforms.
\end{column}
\end{columns}
